Automating G\"odel's Ontological Proof of God's Existence with Higher-order Automated Theorem Provers

### 2014-05-25 Papers/2014-ECAI/ECAI-51.tex
further improvements for final version

\author{Christoph Benzm\"uller\institute{Freie Universit\"at Berlin,
    Germany, email: c.benzmueller@fu-berlin.de}
  \and Bruno Woltzenlogel Paleo\institute{Technical University
    Vienna, Austria,  email: bruno@logic.at}  
}

@@

In theoretical philosophy, formal logical confrontations with such
ontological arguments had been so far (mainly) limited to paper
and pen.  Up to now, the use of computers was prevented, because the
logics of the available theorem proving systems were not expressive
enough to formalize the abstract concepts adequately. G{\"o}del's proof
uses, for example, a complex higher-order modal logic (\HOML) to handle
concepts such as \emph{possible} and \emph{necessary} and to support
quantification over individuals and properties.

Current works \cite{J23,B9} of the first author and Paulson illustrate that many
expressive logics, including quantified (multi-)modal logics, can be
embedded into the classical higher-order logic (\HOL), which can thus be seen
as a universal logic \cite{C36}. For this universal logic, efficient automated
theorem provers have been developed in recent years, and these systems
were now employed in our work.

@@

\begin{definition}\label{holvalue}
  The \emph{value} $\hol{\| s_\alpha\|^{M,g}}$ of a \HOML term
  $\hol{s_\alpha}$ on a model $\hol{M}=\hol{\langle W, R, D, I_{w\in
      W} \rangle}$ in a world $\hol{w}\in \hol{W}$ under assignment $\hol{g}$ is an element $\hol{d}\in \hol{D_\alpha}$
  defined in the following way:
\begin{enumerate}
\item $\hol{\|p_\alpha\|^{M,g}} = \hol{I(p_\alpha)}$ and $\hol{\|X_\alpha\|^{M,g}} = \hol{g(X_\alpha)}$
\item $\hol{\|(s_{\alpha\ar\beta}\, t_\alpha)_\beta\|^{M,g}} = \hol{\|s_{\alpha\ar\beta}\|^{M,g}(\|t_\alpha\|^{M,g})}$
\item $\hol{\|(\lam{X_\alpha}s_\beta)_{\alpha\ar\beta}\|^{M,g}} = $
  the function $\hol{f}$ from $\hol{D_\alpha}$ to $\hol{D_\beta}$ such
  that $\hol{f(d)} = \hol{\|s_\beta\|^{M,g[d/X_\alpha]}}$ for all
  $\hol{d}\in \hol{D_\alpha}$
\item $\hol{\|(\neg_{o\ar o}\, s_o)_o\|^{M,g}} = \hol{T}$ iff $\hol{\|s_o\|^{M,g}} = \hol{F}$
\item $\hol{\|(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)_o\|^{M,g}} =
  \hol{T}$ iff\, $\hol{\|s_o\|^{M,g}} = \hol{T}$ or $\hol{\|t_o\|^{M,g}}
  = \hol{T}$
\item $\hol{\|(\forall_{(\alpha\ar o)\ar o}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{T}$ iff for all $\hol{d}\in
  \hol{D_\alpha}$ we have $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = \hol{T}$
%\item $\|(\Box_{o\ar o}\, s_o)_o\|^{M,g,w} = T$ iff for all $v \in W$ with $w R v$ we have  $\|s_o\|^{M,g,v}  = T$
\end{enumerate}
\end{definition}

@@

\begin{example}\label{ex1} %\footnote{Maybe we should instead use an axiom or theorem from G{\"o}del's proof?}
  The \HOML formula $\Box\,\exi{P_{\mu \ar o}} P\,a_\mu$ is
  associated with the type raised \HOL formula
  $\hol{\llift{\Box}\,\llift{\exists} {P_{\mu\ar
        \sigma}} \lambdot P\ a_\mu}$, which rewrites into the following
  $\beta\eta$-normal \HOL term of type $\sigma$
\[\hol{\lam{W_\iota} \all{V_\iota} \neg (r\,W\,V) \vee \neg \all{P_{\mu\ar \sigma}} \neg (P\, a_\mu\, V)}\]
\end{example}

@@

%\paragraph{Acknowledgments}
\section*{ACKNOWLEDGEMENTS}
 This work has been supported by the German Research Foundation under grants BE 2501/9-1 and BE 2501/11-1. %We thank the anonymous reviewers for their comments and suggestions.

>>>>

\author{Christoph Benzm\"uller\institute{Freie Universit\"at Berlin,
    Germany, email: c.benzmueller@fu-berlin.de; this author has been
    supported by the German National Research Foundation (DFG) under
 grants BE 2501/9-1 and BE 2501/11-1. }
  \and Bruno Woltzenlogel Paleo\institute{Technical University
    Vienna, Austria,  email: bruno@logic.at}  
}

@@

In theoretical philosophy, formal logical confrontations with such
ontological arguments had been so far (mainly) limited to paper
and pen.  Up to now, the use of computers was prevented, because the
logics of the available theorem proving systems were not expressive
enough to formalize the abstract concepts adequately. G{\"o}del's proof
uses, for example, a complex higher-order modal logic (\HOML) to handle
concepts such as \emph{possibility} and \emph{necessity} and to support
quantification over individuals and properties.

Current works \cite{J23,B9} of the first author and Paulson illustrate that many
expressive logics, including quantified (multi-)modal logics, can be
embedded into the classical higher-order logic (\HOL), which can thus be seen
as an universal logic \cite{C36}. For this universal logic, efficient automated
theorem provers have been developed in recent years, and these systems
were now employed in our work.

@@

\begin{definition}\label{holvalue}
  The \emph{value} $\hol{\| s_\alpha\|^{M,g}}$ of a \HOL term
  $\hol{s_\alpha}$ on a model $\hol{M}=\hol{\langle D, I \rangle}$ under assignment $\hol{g}$ is an element $\hol{d}\in \hol{D_\alpha}$
  defined in the following way:
\begin{enumerate}
\item $\hol{\|p_\alpha\|^{M,g}} = \hol{I(p_\alpha)}$ and $\hol{\|X_\alpha\|^{M,g}} = \hol{g(X_\alpha)}$
\item $\hol{\|(s_{\alpha\ar\beta}\, t_\alpha)_\beta\|^{M,g}} = \hol{\|s_{\alpha\ar\beta}\|^{M,g}(\|t_\alpha\|^{M,g})}$
\item $\hol{\|(\lam{X_\alpha}s_\beta)_{\alpha\ar\beta}\|^{M,g}} = $
  the function $\hol{f}$ from $\hol{D_\alpha}$ to $\hol{D_\beta}$ such
  that $\hol{f(d)} = \hol{\|s_\beta\|^{M,g[d/X_\alpha]}}$ for all
  $\hol{d}\in \hol{D_\alpha}$
\item $\hol{\|(\neg_{o\ar o}\, s_o)_o\|^{M,g}} = \hol{T}$ iff $\hol{\|s_o\|^{M,g}} = \hol{F}$
\item $\hol{\|(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)_o\|^{M,g}} =
  \hol{T}$ iff\, $\hol{\|s_o\|^{M,g}} = \hol{T}$ or $\hol{\|t_o\|^{M,g}}
  = \hol{T}$
\item $\hol{\|(\forall_{(\alpha\ar o)\ar o}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{T}$ iff for all $\hol{d}\in
  \hol{D_\alpha}$ we have $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = \hol{T}$
%\item $\|(\Box_{o\ar o}\, s_o)_o\|^{M,g,w} = T$ iff for all $v \in W$ with $w R v$ we have  $\|s_o\|^{M,g,v}  = T$
\end{enumerate}
\end{definition}

@@

\begin{example}\label{ex1} %\footnote{Maybe we should instead use an axiom or theorem from G{\"o}del's proof?}
  The \HOML formula $\Box\,\exi{P_{\mu \ar o}} P\,a_\mu$ is
  associated with the type raised \HOL formula
  $\hol{\llift{\Box}\,\llift{\exists} {P_{\mu\ar
        \sigma}} \lambdot P\ a_\mu}$, which rewrites into the following
  $\beta\eta$-normal \HOL term of type $\hol{\sigma}$
\[\hol{\lam{W_\iota} \all{V_\iota} \neg (r\,W\,V) \vee \neg \all{P_{\mu\ar \sigma}} \neg (P\, a_\mu\, V)}\]
\end{example}

@@

%\paragraph{Acknowledgments}
\section*{ACKNOWLEDGEMENTS}
 We thank Alexander Steen,
 Max Wisniewski, and the anonymous reviewers for their comments and suggestions.
